Quotients of the magmatic operad:\\ lattice structures and convergent rewrite systems

### 2018-09-11 Mag_quotients.tex
MAJ

\Todo{S: ce thm est une "série génératriciation" du lemme précédent. Je
serais d'avis de supprimer ce lemme et de montrer directement le thm.
Avec l'espace gagné, je propose de donner pour cette partie un exemple:
deux quotients --- linéaires --- de Mag dont on donne les séries
de Hilbert (quitte à ne pas les démontrer, mais pour illustrer le thm).}

>>>>

We terminate this section with an example illustrating the previous
constructions in the case where the characteristic of the
field $\K$ is different from $2$. For that, we introduce various operads,
which requires the following notations, also used in
Section~\ref{sec:CAs_d}. Let, for any integer $\gamma \geq 1$, the binary
trees $\LComb{\gamma}$ and $\RComb{\gamma}$ be respectively the left and
the right combs of degree $\gamma$. These trees are depicted as
\begin{equation}
    \LComb{\gamma} = \enspace
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-5.25){};
        \node(2)at(2.00,-5.25){};
        \node(4)at(4.00,-3.50){};
        \node(6)at(6.00,-1.75){};
        \node[NodeST](1)at(1.00,-3.50){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,0.00){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge,dotted](1)edge[]node[font=\tiny]{
            \begin{math}\gamma\! -\! 1\end{math}\hspace*{.6cm}}(3);
        \draw[Edge](2)--(1);
        \draw[Edge](3)--(5);
        \draw[Edge](4)--(3);
        \draw[Edge](6)--(5);
        \node(r)at(5.00,1.5){};
        \draw[Edge](r)--(5);
    \end{tikzpicture}
    \qquad \mbox{and} \qquad
    \RComb{\gamma} =
    \begin{tikzpicture}[xscale=.26,yscale=.3,Centering]
        \node(0)at(0.00,-1.75){};
        \node(2)at(2.00,-3.50){};
        \node(4)at(4.00,-5.25){};
        \node(6)at(6.00,-5.25){};
        \node[NodeST](1)at(1.00,0.00){\begin{math}\Product\end{math}};
        \node[NodeST](3)at(3.00,-1.75){\begin{math}\Product\end{math}};
        \node[NodeST](5)at(5.00,-3.50){\begin{math}\Product\end{math}};
        \draw[Edge](0)--(1);
        \draw[Edge](2)--(3);
        \draw[Edge](3)--(1);
        \draw[Edge](4)--(5);
        \draw[Edge,dotted](5)edge[]node[font=\tiny]{
            \hspace*{.6cm}\begin{math}\gamma\! -\! 1\end{math}}(3);
        \draw[Edge](6)--(5);
        \node(r)at(1.00,1.5){};
        \draw[Edge](r)--(1);
    \end{tikzpicture}\,,
\end{equation}
where the values on the dotted edges denote the number of internal
nodes they contain.

We define the \Def{anti-associative operad} by
\begin{equation}
  \label{equ:anti-associative_operad}
  \AAs:=\KMag/_{\IAAs}
\end{equation}
where $\IAAs$ is the ideal spanned by
\begin{math}
  \LComb{3}+\RComb{3}
\end{math}.
\Todo{Illustrer avec la représentation graphique ?}

We define the \Def{2-nihilpotente operad} by
\begin{equation}
  \label{equ:2-nihilpotente_operad}
  \NiHil:=\KMag/_{\INiHil}
  \end{equation}
where $\INiHil$ is the ideal spanned by the two trees $\LComb{3}$ and
$\RComb{3}$.
\Todo{Est-ce qu'il faut rajouter le 2 dans la macro désignant cette
  opérade ? De plus, Samuele,\\
  pourrais-tu rajouter la référence de Loday ?}
We easily check that we have
\begin{equation}
  \label{equ:Hilbert_series_of_NiHil}
  \HilbertSeries_{\NiHil}(t)=t+t^2.
\end{equation}
\Todo{Est-ce bien la bonne série ? Samuele, tu avais l'air de dire que
  c'était plutôt $t+2t^2$}

Finally, for every integer $\gamma\geq 2$, we define the
\Def{$\gamma$-right comb operad} $\RC{\gamma}$ as follows. For every
arity $n$, we let
\begin{equation}
  \RC{\gamma}(n):=
    \begin{cases}
        \As(n) &
        \mbox{if } n \leq \gamma-1, \\
        \RComb{n}
            & \mbox{otherwise},
    \end{cases}
\end{equation}
and the partial composition $\Tfr_1 \circ_i \Tfr_2$ is the partial
composition of $\Tfr_1$ and $\Tfr_2$ in $\Mag$ if the integer
\begin{math}
  n:=|\Tfr_1|+|\Tfr_2| 
\end{math}
is strictly smaller than $\gamma$, and $\RComb{n}$, otherwise.
\medbreak

\begin{Lemma}
  \label{lem:presentation_of_RC}
  We have
  \begin{equation}
  \label{equ:presentation_of_RC}
  \NiHil:=\KMag/_{\IRCC{\gamma}}
  \end{equation}
  where $\IRC{\gamma}$ is the ideal spanned by the elements
  $\Tfr-\RComb{\gamma}$, where $\Tfr$ runs over all the trees of arity
  $\gamma$.
\end{Lemma}

\begin{proof}
  \Todo{todo}
\end{proof}

\medbreak
The lower-bound and the upper-bound of $\As$ and $\AAs$ are given by the
following.
\begin{Theorem}
  \label{thm:example_lattice}
  We have
  \begin{equation}
    \label{equ:example_lower-bound}
    \As\InfQMag\AAs=\NiHil
  \end{equation}
  and
  \begin{equation}
    \label{equ:example_upper-bound}
    \As\SupQMag\AAs=\RC{3}.
    \end{equation}
  \end{Theorem}

\begin{proof}
  First, we show~\eqref{equ:example_lower-bound}. By definition of the
  lower-bound $\InfQMag$, the ideal of relations of $\As\InfQMag\AAs$
  is spanned by the two elements $\LComb{3}-\RComb{3}$ and
  $\LComb{3}+\RComb{3}$. By linear transformations applied to these
  generators, the ideal of relations of $\As\InfQMag\AAs$ is spanned by
  $\LComb{3}$ and $\RComb{3}$, that is it is equal to $\INiHil$, which
  proves~\eqref{equ:example_lower-bound}.

\smallbreak
Now, we show~\eqref{equ:example_upper-bound}. We denote by
\begin{math}
  \pi:\KMag\to\RC{3}
\end{math}
the natural projection. Let $\Tfr$ be a tree of arity $3$. Then,
$\Tfr-\RComb{\gamma}$ belongs to the ideals $\IAs$ and $\IAAs$ of
relations of $\As$ and $\AAs$, respectively. Indeed, every
tree of arity $3$ to the same element $\RComb{\gamma}$. Hence, by
definition of the upper-bound $\SupQMag$, $\pi$ induces a surjective
operad morphism
\begin{math}
  \overline{\pi}:\As\SupQMag\AAs\to\RC{3}
\end{math}.
We conclude using a dimension argument: by definition, the Hilbert series
of $\RC{3}$ is equal to
\begin{equation}
  \label{equ:Hilbert_series_of_RC3}
  t+t^2+2t^3+\sum_{n\geq 4}t^n
\end{equation}
and from the Grassmann formula analog given in
Theorem~\ref{thm:Grassmann_formula_for_Hilbert_series_of_QMag}, we check
that the Hilbert series of $\As\SupQMag\AAs$ is also equal
to~\eqref{equ:Hilbert_series_of_RC3}. Hence, $\overline{\pi}$ is an
isomorphism, which proves~\eqref{equ:example_upper-bound}.
  \end{proof}
\medbreak
